---
title: "Analysis and Study of UCI Heart Disease dataset"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r, echo=FALSE, comment="", tidy=TRUE, error=TRUE}
set.seed(1234)
today <- date()
today <- unlist(strsplit(today, " "))
cat("by Prateek Sarangi,", today[1], ",", today[2:3], today[5])
```

Part 1:- Getting the feature dataset from the actual data
---------------------------------------------------------------------------
## Dataset details
### Attribute information
- age  
- sex  
- chest pain type (4 values) -> cp  
- resting blood pressure -> trestbps  
- serum cholestoral in mg/dl-> chol  
- fasting blood sugar > 120 mg/dl -> fbs  
- resting electrocardiographic results (values 0,1,2) -> restecg  
- maximum heart rate achieved -> thalach  
- exercise induced angina -> exang  
- oldpeak = ST depression induced by exercise relative to rest -> oldpeak  
- the slope of the peak exercise ST segment -> slope  
- number of major vessels (0-3) colored by flourosopy -> ca  
- thal: 3 = normal; 6 = fixed defect; 7 = reversable defect  

## Program for exploratory analysis and preprocessing
### Function to Normalize the data
```{r}
normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}
```

### Import dataset and view it as a dataframe
```{r}
heart <- read.csv("~/HeartDisease/heart.csv")
head(heart)
```

### Normalize the age and separate it in various groups to show which age group is more prone to heart disease 
It uses the normalize function defined above and then categorize the data into **four** groups.  
After normalization the valuesaranges from **0-1**  
It is grouped as follows  
- **0.00 to 0.25** as **Group 1** giving it value **0.1**  
- **0.25 to 0.50** as **Group 2** giving it value **0.4**  
- **0.50 to 0.75** as **Group 3** giving it value **0.6**  
- **0.75 to 1.00** as **Group 4** giving it value **0.9**  
```{r}
library(dplyr)
dfNorm <- as.data.frame(lapply(heart["age"], normalize))
heart["age"] <- dfNorm
heart["age"] <- as.data.frame(lapply(heart["age"], function(x){replace(x,between(x, 0.0, 0.25), 0.1)}))
heart["age"] <- as.data.frame(lapply(heart["age"], function(x){replace(x,between(x, 0.25, 0.6), 0.4)}))
heart["age"] <- as.data.frame(lapply(heart["age"], function(x){replace(x,between(x, 0.5, 0.75), 0.6)}))
heart["age"] <- as.data.frame(lapply(heart["age"], function(x){replace(x,between(x, 0.75, 1), 0.9)}))
```

### Normalize the Rest Blood Pressure and separate it in various groups to show what blood pressure group is more prone to heart disease 
It uses the normalize function defined above and then categorize the data into **three** groups.  
After normalization the valuesaranges from **0-1**  
It is grouped as follows  
- **0.00 to 0.33** as **Group 1** giving it value **0.2**  
- **0.33 to 0.67** as **Group 2** giving it value **0.6**  
- **0.67 to 1.00** as **Group 3** giving it value **1.0**  
```{r}
dfNorm <- as.data.frame(lapply(heart["trestbps"], normalize))
heart["trestbps"] <- dfNorm
heart["trestbps"] <- as.data.frame(lapply(heart["trestbps"], function(x){replace(x, between(x, 0.0, 0.33), 0.2)}))
heart["trestbps"] <- as.data.frame(lapply(heart["trestbps"], function(x){replace(x, between(x, 0.33, 0.67), 0.6)}))
heart["trestbps"] <- as.data.frame(lapply(heart["trestbps"], function(x){replace(x, between(x, 0.67, 1), 1)}))
```

### Normalize the Cholestrole level and separate it in various groups to show what cholestrole level is more prone to heart disease 
It uses the normalize function defined above and then categorize the data into **five** groups.  
After normalization the valuesaranges from **0-1**  
It is grouped as follows  
- **0.00 to 0.20** as **Group 1** giving it value **0.1**  
- **0.20 to 0.40** as **Group 2** giving it value **0.3**  
- **0.40 to 0.60** as **Group 3** giving it value **0.5**  
- **0.60 to 0.80** as **Group 4** giving it value **0.7**  
- **0.80 to 1.00** as **Group 5** giving it value **0.9**  
```{r}
dfNorm <- as.data.frame(lapply(heart["chol"], normalize))
heart["chol"] <- dfNorm
heart["chol"] <- as.data.frame(lapply(heart["chol"], function(x){replace(x, between(x, 0.0, 0.2), 0.1)}))
heart["chol"] <- as.data.frame(lapply(heart["chol"], function(x){replace(x, between(x, 0.2, 0.4), 0.3)}))
heart["chol"] <- as.data.frame(lapply(heart["chol"], function(x){replace(x, between(x, 0.4, 0.6), 0.5)}))
heart["chol"] <- as.data.frame(lapply(heart["chol"], function(x){replace(x, between(x, 0.6, 0.8), 0.7)}))
heart["chol"] <- as.data.frame(lapply(heart["chol"], function(x){replace(x, between(x, 0.8, 1), 0.9)}))
```

### Normalize the chest pain and separate it in various groups to show what percentage of chest pain is more prone to heart disease 
It uses the normalize function defined above and then categorize the data into **four** groups.  
After normalization the valuesaranges from **0-1**  
It is grouped as follows  
- **0.00 to 0.25** as **Group 1** giving it value **0.1**  
- **0.25 to 0.50** as **Group 2** giving it value **0.4**  
- **0.50 to 0.75** as **Group 3** giving it value **0.6**  
- **0.75 to 1.00** as **Group 4** giving it value **0.9**  
```{r}
dfNorm <- as.data.frame(lapply(heart["cp"], normalize))
heart["cp"] <- dfNorm
heart["cp"] <- as.data.frame(lapply(heart["cp"], function(x){replace(x, between(x, 0.0, 0.25), 0.1)}))
heart["cp"] <- as.data.frame(lapply(heart["cp"], function(x){replace(x, between(x, 0.25, 0.5), 0.4)}))
heart["cp"] <- as.data.frame(lapply(heart["cp"], function(x){replace(x, between(x, 0.5, 0.75), 0.6)}))
heart["cp"] <- as.data.frame(lapply(heart["cp"], function(x){replace(x, between(x, 0.75, 1), 0.9)}))
```

### Normailze the maximum heart rate of the patient
```{r}
dfNorm <- as.data.frame(lapply(heart["thalach"], normalize))
heart["thalach"] <- dfNorm
```

### Normalize the thal and separate it in various groups to show what thal is more prone to heart disease 
It uses the normalize function defined above and then categorize the data into **three** groups.  
Thal: 3 = normal; 6 = fixed defect; 7 = reversable defect  
After normalization the valuesaranges from **0-1**  
It is grouped as follows  
- **0.00 to 0.33** as **Group 1** giving it value **0.2**  
- **0.33 to 0.67** as **Group 2** giving it value **0.6**  
- **0.67 to 1.00** as **Group 3** giving it value **0.9**  
```{r}
dfNorm <- as.data.frame(lapply(heart["thal"], normalize))
heart["thal"] <- dfNorm
heart["thal"] <- as.data.frame(lapply(heart["thal"], function(x){replace(x, between(x, 0.0, 0.33), 0.2)}))
heart["thal"] <- as.data.frame(lapply(heart["thal"], function(x){replace(x, between(x, 0.33, 0.67), 0.6)}))
heart["thal"] <- as.data.frame(lapply(heart["thal"], function(x){replace(x, between(x, 0.67, 1), 0.9)}))
```

### Normailze the peak exercise ST segment of the patient
```{r}
dfNorm <- as.data.frame(lapply(heart["slope"], normalize))
heart["slope"] <- dfNorm
```

### Normailze the major vessels (0-3) colored by flourosopy
```{r}
dfNorm <- as.data.frame(lapply(heart["ca"], normalize))
heart["ca"] <- dfNorm
```

### Value modification for feeding into Neural Network
Values of **Sex**, **Fasting blood suger**, **Resting electrocardiographic results** and **Exercise induced angina** are replaced so that it can be fed to the neural network.  
- *1* is replaced with *0.9*  
- *0* is replaced with *0.1*  
```{r}
heart["sex"] <- as.data.frame(lapply(heart["sex"], function(x){replace(x, x == 0, 0.1)}))
heart["sex"] <- as.data.frame(lapply(heart["sex"], function(x){replace(x, x == 1, 0.9)}))

heart["fbs"] <- as.data.frame(lapply(heart["fbs"], function(x){replace(x, x == 0, 0.1)}))
heart["fbs"] <- as.data.frame(lapply(heart["fbs"], function(x){replace(x, x == 1, 0.9)}))

heart["restecg"] <- as.data.frame(lapply(heart["restecg"], function(x){replace(x, x == 0, 0.1)}))
heart["restecg"] <- as.data.frame(lapply(heart["restecg"], function(x){replace(x, x == 1, 0.9)}))

heart["exang"] <- as.data.frame(lapply(heart["exang"], function(x){replace(x, x == 0, 0.1)}))
heart["exang"] <- as.data.frame(lapply(heart["exang"], function(x){replace(x, x == 1, 0.9)}))
```

## The dataset after the modifications are made.
```{r echo = FALSE}
head(heart)
```

## Splitting of training and testing data
We are using both randomly generated and sequentially choosen 75% of the data as training set and rest 25% as our test set.  
**train_ind_rand** gives the indeces of the samples which are to be used as the training sample in the dataset.  
**trainrand** gives the randomly choosen train dataset.  
**testrand** given the randomly choosen test dataset.  
**trainseq** gives the sequencially choosen train dataset.  
**testseq** gives the sequentially choosen test dataset.  
```{r}
smp_size <- floor(0.75 * nrow(heart))
train_ind_rand <- sample(seq_len(nrow(heart)), size = smp_size)

trainrand <- heart[train_ind_rand, ]
testrand <- heart[-train_ind_rand, ]

trainseq <- heart[1:227, ]
testseq <- heart[227:303, ]
```

## Writting the dataset into csv files so that it can be used in the Python program for Neural network classification.
- **heart1.csv** contain the modified Heart Disease dataset.  
- **trainrand.csv** conains the randomly chosen train dataset.  
- **testrand.csv** conains the randomly chosen test dataset.  
- **trainseq.csv** conains the sequencially choosen train dataset.  
- **testseq.csv** conains the sequentially choosen test dataset. 
```{r}
write.csv(heart, "~/HeartDisease/heart1.csv", row.names = FALSE)
write.csv(trainrand, "~/HeartDisease/trainrand.csv", row.names = FALSE)
write.csv(testrand, "~/HeartDisease/testrand.csv", row.names = FALSE)
write.csv(trainseq, "~/HeartDisease/trainseq.csv", row.names = FALSE)
write.csv(testseq, "~/HeartDisease/testseq.csv", row.names = FALSE)
```

Part 2:- Analysis of the neural network model
---------------------------------------------------------------------------
## Plot for varying the number of hidden layers
Neural network with various number of hidden layers are tested and the results of the test accuracy are ploted.  
Input layer contains **Thirteen** input neurons, each hidden layer has **Tweleve** neurons and the output layer has **One** neuron with **Two** classes.  
- Class zero -> The patient is not suffering from heart disease.  
- Class one -> The patient is suffering from heart disease.  

### Plot of test accuracy for different hidden layers
```{r echo=FALSE}
LayerAcc <- read.csv("~/HeartDisease/LayerAcc.csv", header=FALSE)
LayerAcc.bar <- barplot(LayerAcc$V2, names.arg=LayerAcc$V1, col = 'blue')
```

### Conclusion from the plot
According to the test runs we can see that model with **Three**, **Four**, **Five** and **Six** hidden layers, having **Tweleve** neuron each perform better than the other choosen models.  

## Plot for varying the number of neurons
## Plots for Accuracy and Mean Square Error
The neural network with the best results is run **Twinty** times and the average mean square error, and accuracy was taken and plot in the following graph.  
```{r}
```

### Cloclusion from the plot


### Structure of the model
- Input layer -> 13 input neuron  
- First hidden layer -> 12 hidden neuraon  
- Second hidden layer -> 12 hidden neuraon  
- Third hidden layer -> 10 hidden neuraon  
- Fourth hidden layer -> 8 hidden neuraon  
- Fifth hidden layer -> 8 hidden neuraon  
- Output layer -> 1 neuron, 2 classes  

### Plot for Accuracy of the model
```{r echo = FALSE}
tableAccuracy <- read.csv("~/HeartDisease/tableAccuracy.csv", header=FALSE, sep=";")

meanAccuracy <- colMeans(tableAccuracy[sapply(tableAccuracy, is.numeric)])
itr <- 1:1500

plot(itr, meanAccuracy, "l", xlab = "Iteration", ylab = "Accuracy",
     col = "blue", lwd = 2)
```

### Plot for Mean Square Error of the model
```{r echo = FALSE}
tableMSE <- read.csv("~/HeartDisease/tableMSE.csv", header=FALSE, sep=";")

meanMSE <- colMeans(tableMSE[sapply(tableMSE, is.numeric)])
itr <- 1:1500

plot(itr, meanMSE, "l", xlab = "Iteration", ylab = "Mean Square Error",
     col = "orange", lwd = 2)
```






