---
title: "Heart Disease UCI, Analysis and Study"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Analysis and Study of UCI Heart Disease dataset

```{r, echo=FALSE, comment="", tidy=TRUE, error=TRUE}
set.seed(1234)
today <- date()
today <- unlist(strsplit(today, " "))
cat("by Prateek Sarangi,", today[1], ",", today[2:3], today[5])
```

## Dataset details
### Attribute information
- age  
- sex  
- chest pain type (4 values) -> cp  
- resting blood pressure -> trestbps  
- serum cholestoral in mg/dl-> chol  
- fasting blood sugar > 120 mg/dl -> fbs  
- resting electrocardiographic results (values 0,1,2) -> restecg  
- maximum heart rate achieved -> thalach  
- exercise induced angina -> exang  
- oldpeak = ST depression induced by exercise relative to rest -> oldpeak  
- the slope of the peak exercise ST segment -> slope  
- number of major vessels (0-3) colored by flourosopy -> ca  
- thal: 3 = normal; 6 = fixed defect; 7 = reversable defect  

## Program for exploratory analysis and preprocessing
### Function to Normalize the data
```{r}
normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}
```

### Import dataset and view it as a dataframe
```{r}
heart <- read.csv("~/HeartDisease/heart.csv")
head(heart)
```

### Normalize the age and separate it in various groups to show which age group is more prone to heart disease 
It uses the normalize function defined above and then categorize the data into **four** groups.  
After normalization the valuesaranges from **0-1**  
It is grouped as follows  
- **0.00 to 0.25** as **Group 1** giving it value **0.1**  
- **0.25 to 0.50** as **Group 2** giving it value **0.4**  
- **0.50 to 0.75** as **Group 3** giving it value **0.6**  
- **0.75 to 1.00** as **Group 4** giving it value **0.9**  
```{r}
library(dplyr)
dfNorm <- as.data.frame(lapply(heart["age"], normalize))
heart["age"] <- dfNorm
heart["age"] <- as.data.frame(lapply(heart["age"], function(x){replace(x,between(x, 0.0, 0.25), 0.1)}))
heart["age"] <- as.data.frame(lapply(heart["age"], function(x){replace(x,between(x, 0.25, 0.6), 0.4)}))
heart["age"] <- as.data.frame(lapply(heart["age"], function(x){replace(x,between(x, 0.5, 0.75), 0.6)}))
heart["age"] <- as.data.frame(lapply(heart["age"], function(x){replace(x,between(x, 0.75, 1), 0.9)}))
```

### Normalize the Rest Blood Pressure and separate it in various groups to show what blood pressure group is more prone to heart disease 
It uses the normalize function defined above and then categorize the data into **three** groups.  
After normalization the valuesaranges from **0-1**  
It is grouped as follows  
- **0.00 to 0.33** as **Group 1** giving it value **0.2**  
- **0.33 to 0.67** as **Group 2** giving it value **0.6**  
- **0.67 to 1.00** as **Group 3** giving it value **1.0**  
```{r}
dfNorm <- as.data.frame(lapply(heart["trestbps"], normalize))
heart["trestbps"] <- dfNorm
heart["trestbps"] <- as.data.frame(lapply(heart["trestbps"], function(x){replace(x, between(x, 0.0, 0.33), 0.2)}))
heart["trestbps"] <- as.data.frame(lapply(heart["trestbps"], function(x){replace(x, between(x, 0.33, 0.67), 0.6)}))
heart["trestbps"] <- as.data.frame(lapply(heart["trestbps"], function(x){replace(x, between(x, 0.67, 1), 1)}))
```

### Normalize the Cholestrole level and separate it in various groups to show what cholestrole level is more prone to heart disease 
It uses the normalize function defined above and then categorize the data into **five** groups.  
After normalization the valuesaranges from **0-1**  
It is grouped as follows  
- **0.00 to 0.20** as **Group 1** giving it value **0.1**  
- **0.20 to 0.40** as **Group 2** giving it value **0.3**  
- **0.40 to 0.60** as **Group 3** giving it value **0.5**  
- **0.60 to 0.80** as **Group 4** giving it value **0.7**  
- **0.80 to 1.00** as **Group 5** giving it value **0.9**  
```{r}
dfNorm <- as.data.frame(lapply(heart["chol"], normalize))
heart["chol"] <- dfNorm
heart["chol"] <- as.data.frame(lapply(heart["chol"], function(x){replace(x, between(x, 0.0, 0.2), 0.1)}))
heart["chol"] <- as.data.frame(lapply(heart["chol"], function(x){replace(x, between(x, 0.2, 0.4), 0.3)}))
heart["chol"] <- as.data.frame(lapply(heart["chol"], function(x){replace(x, between(x, 0.4, 0.6), 0.5)}))
heart["chol"] <- as.data.frame(lapply(heart["chol"], function(x){replace(x, between(x, 0.6, 0.8), 0.7)}))
heart["chol"] <- as.data.frame(lapply(heart["chol"], function(x){replace(x, between(x, 0.8, 1), 0.9)}))
```

### Normalize the chest pain and separate it in various groups to show what percentage of chest pain is more prone to heart disease 
It uses the normalize function defined above and then categorize the data into **four** groups.  
After normalization the valuesaranges from **0-1**  
It is grouped as follows  
- **0.00 to 0.25** as **Group 1** giving it value **0.1**  
- **0.25 to 0.50** as **Group 2** giving it value **0.4**  
- **0.50 to 0.75** as **Group 3** giving it value **0.6**  
- **0.75 to 1.00** as **Group 4** giving it value **0.9**  
```{r}
dfNorm <- as.data.frame(lapply(heart["cp"], normalize))
heart["cp"] <- dfNorm
heart["cp"] <- as.data.frame(lapply(heart["cp"], function(x){replace(x, between(x, 0.0, 0.25), 0.1)}))
heart["cp"] <- as.data.frame(lapply(heart["cp"], function(x){replace(x, between(x, 0.25, 0.5), 0.4)}))
heart["cp"] <- as.data.frame(lapply(heart["cp"], function(x){replace(x, between(x, 0.5, 0.75), 0.6)}))
heart["cp"] <- as.data.frame(lapply(heart["cp"], function(x){replace(x, between(x, 0.75, 1), 0.9)}))
```

### Normailze the maximum heart rate of the patient
```{r}
dfNorm <- as.data.frame(lapply(heart["thalach"], normalize))
heart["thalach"] <- dfNorm
```

### Normalize the thal and separate it in various groups to show what thal is more prone to heart disease 
It uses the normalize function defined above and then categorize the data into **three** groups.  
Thal: 3 = normal; 6 = fixed defect; 7 = reversable defect  
After normalization the valuesaranges from **0-1**  
It is grouped as follows  
- **0.00 to 0.33** as **Group 1** giving it value **0.2**  
- **0.33 to 0.67** as **Group 2** giving it value **0.6**  
- **0.67 to 1.00** as **Group 3** giving it value **0.9**  
```{r}
dfNorm <- as.data.frame(lapply(heart["thal"], normalize))
heart["thal"] <- dfNorm
heart["thal"] <- as.data.frame(lapply(heart["thal"], function(x){replace(x, between(x, 0.0, 0.33), 0.2)}))
heart["thal"] <- as.data.frame(lapply(heart["thal"], function(x){replace(x, between(x, 0.33, 0.67), 0.6)}))
heart["thal"] <- as.data.frame(lapply(heart["thal"], function(x){replace(x, between(x, 0.67, 1), 0.9)}))
```

### Normailze the peak exercise ST segment of the patient
```{r}
dfNorm <- as.data.frame(lapply(heart["slope"], normalize))
heart["slope"] <- dfNorm
```

### Normailze the major vessels (0-3) colored by flourosopy
```{r}
dfNorm <- as.data.frame(lapply(heart["ca"], normalize))
heart["ca"] <- dfNorm
```

### Value modification for feeding into Neural Network
Values of **Sex**, **Fasting blood suger**, **Resting electrocardiographic results** and **Exercise induced angina** are replaced so that it can be fed to the neural network.  
- *1* is replaced with *0.9*  
- *0* is replaced with *0.1*  
```{r}
heart["sex"] <- as.data.frame(lapply(heart["sex"], function(x){replace(x, x == 0, 0.1)}))
heart["sex"] <- as.data.frame(lapply(heart["sex"], function(x){replace(x, x == 1, 0.9)}))

heart["fbs"] <- as.data.frame(lapply(heart["fbs"], function(x){replace(x, x == 0, 0.1)}))
heart["fbs"] <- as.data.frame(lapply(heart["fbs"], function(x){replace(x, x == 1, 0.9)}))

heart["restecg"] <- as.data.frame(lapply(heart["restecg"], function(x){replace(x, x == 0, 0.1)}))
heart["restecg"] <- as.data.frame(lapply(heart["restecg"], function(x){replace(x, x == 1, 0.9)}))

heart["exang"] <- as.data.frame(lapply(heart["exang"], function(x){replace(x, x == 0, 0.1)}))
heart["exang"] <- as.data.frame(lapply(heart["exang"], function(x){replace(x, x == 1, 0.9)}))
```

## The dataset after the modifications are made.
```{r eval=TRUE, epoch = FALSE}
head(heart)
```

## Splitting of training and testing data
We are using both randomly generated and sequentially choosen 75% of the data as training set and rest 25% as our test set.  
**train_ind_rand** gives the indeces of the samples which are to be used as the training sample in the dataset.  
**trainrand** gives the randomly choosen train dataset.  
**testrand** given the randomly choosen test dataset.  
**trainseq** gives the sequencially choosen train dataset.  
**testseq** gives the sequentially choosen test dataset.  
```{r}
smp_size <- floor(0.75 * nrow(heart))
train_ind_rand <- sample(seq_len(nrow(heart)), size = smp_size)

trainrand <- heart[train_ind_rand, ]
testrand <- heart[-train_ind_rand, ]

trainseq <- heart[1:227, ]
testseq <- heart[227:303, ]
```

## Writting the dataset into csv files so that it can be used in the Python program for Neural network classification.
- **heart1.csv** contain the modified Heart Disease dataset.  
- **trainrand.csv** conains the randomly chosen train dataset.  
- **testrand.csv** conains the randomly chosen test dataset.  
- **trainseq.csv** conains the sequencially choosen train dataset.  
- **testseq.csv** conains the sequentially choosen test dataset. 
```{r}
write.csv(heart, "~/HeartDisease/heart1.csv", row.names = FALSE)
write.csv(trainrand, "~/HeartDisease/trainrand.csv", row.names = FALSE)
write.csv(testrand, "~/HeartDisease/testrand.csv", row.names = FALSE)
write.csv(trainseq, "~/HeartDisease/trainseq.csv", row.names = FALSE)
write.csv(testseq, "~/HeartDisease/testseq.csv", row.names = FALSE)
```









